---
title: "Extinction and origination dynamics in Tr-J"
author: "ALLuza, MG Bender, CS Dambros, L Kerber - Departamento de Ecologia e Evolução, Universidade Federal de Santa Maria"
date: "2023-03-23"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- badges: start -->
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 1. Introduction

#### A bit about diversification and the fossil record
Extinctions are large depressions on Earth biodiversity history (Benton 1995). The diversity of all organisms on Earth, with an likely origin during the Precambrian period (4560 to 570 Ma), has increased rapidly and substantially after the Vendian and Early Cambrian (600 Ma) (Benton 1995). Since then, periods of diversification (i.e. the balance of extinctions and speciations/originations) were marked by periods of high rates of origination interrupted/interpassed by periods of biodiversity crisis/depression, called 'mass extinctions' (Benton 1995). The fossil record supports five mass extinction events, known as the 'big five': end Permian, late Ordovician, late Devonian, Triassic-Jurassic boundary, and Cretaceous-Paleogene boundary -- that  vanished from Earth a considerable proportion of biodiversity, while speciation and radiation to vacant niches occurred for remaining taxa from there on (Benton 1995, May 1997, Reeves et al. 2020).
  
The Triassic-Jurassic extinction (Rhaetian-Hettangian intervals) played an important role in the diversification rates of archosaurs, therapsid reptiles, ammonites, among others (Allen et al. 2018), yet recent research cast doubt on their existence (Bambach et al. 2004). The statistical support for this extinction event (and its place as one of the 'big five' extinctions) is limited due to the generally small sample size at the Rhaetian-Hettangian intervals. Existing evidence suggests that the lower number of taxa at this period can be a byproduct of high extinction rates (e.g., Wignall & Atkinson 2020; Olsen et al. 2022) --which seemed to exert an impressive effect on marine organisms (Benton 1995)-- as well the low origination/speciation rates and nearly constant extinction rates compared to previous time (Bamchach et al. 2004). Thus, models that account for both uneven sampling size across periods and the state of taxa occurrence at previous time could provide a more robust inference on/shed light on the processes that caused the Triassic-Jurassic extinction, whether low speciation/low extinction rates or high extinction/low speciation rates.
    
#### The drivers of distribution
Understanding what drives shifts in temporal and spatial distribution of taxa is one of the major goals in biology, disregarding whether the studied taxa are extinct or extant. In the case of extinct taxa, the abiotic environment and its variation over millennia years might affect extinction and origination rates (e.g., Fritz et al. 2013), while factors that hamper data collection and the taphonomic process (e.g., sedimentation, transport) might bias species detection in fossilized samples (Liow 2013). The simultaneous accounting of ecological and sampling processes yielding modeled data in hierarchical models has some tradition in ecology (Williams et al. 2002, McKenzie et al. 2002, Kery et al. 2013, Guillera-Arroita et al. 2016; reviewed by Kellner & Swihart 2014) but the use of these models to make inference in paleontology is still in its infancy (Liow 2013, Lawing et al. 2021). Accounting for these drivers of distribution might provide better maps, more robust inferences and treatment of biological and observational uncertainty producing paleontological data (Liow 2013 for a paleontological perspective; Kery et al. 2013, Guillera-Arroita et al. 2016 show ecological (neontological) perspectives).
    
#### Major aim
Here we used ~2 million fossil observations comprising 1139 genera and 1770 geologic formations, collected across the entire Triassic-Jurassic periods, to estimate pairwise rates of extinction and origination across 17 time intervals of the Triassic and Jurassic periods (Olenekian (early Triassic) to Tithonian (late Jurassic)). We emphasized i) temporal variation on these rates, ii) the proportional change in number of genus from the beggining of each interval to its end (Bambach et al., 2004), and iii) the expected number of genera across these intervals. As the limit between the Rhaetian (age) and Hettangian (age) marked the Triassic-Jurassic extinction event and the split between these periods, we expected to find i) higher rates of extinction and lower rates of origination, ii) high change in the number of genus from the beggining to the end of these periods, and iii) a lower expected number of taxa at Rhaetian-Hettangian intervals when compared to other intervals in the data.    


# Methods
We used the Paleo Database (REF) as the main source of paleontological data for the Triassic-Jurassic period. We used information at the genus level, as genera give a closer view of species-level behavior through time, despite comprising less complete data than one at a higher taxonomic level because of the vagaries of preservation, collection, and study (Benton 1995). Nonetheless, we accounted for such sampling biases in our models.

## State-space models

State-space models are hierarchical models that enable the estimation of biological parameters describing dynamic properties of ecological systems (Kéry & Royle 2007, Ecology). They are called 'first-order Markovian models' because they use information from previous time and/or neighbor sites to project biological parameters in the next time. Thus, they are useful to estimating parameters involved with inference in paleontology, such as extinction (*epslon*), origination (*gamma*), and turnover rates (Bambach et al. 2004, Benton 1995). In state-space models, the true state of a system (e.g. the realized occurrence of the fossil genera *g* at the time interval *t*) is denoted by the latent, stochastic variable *z~[t,g]*.  The *z~[t,g]* is the realization of a Bernoulli process ('dbern') dependent on the variable *muZ~[t,g]*, which is the expected probability of occupancy of interval *t* by the genus *g*. In its turn, the *muZ~[t,g]* depends on both the realized occurrence of genus *g* in the previous time *t-1*, ie. *z~[t-1,g]* (whose first time gives the initial conditions for the dynamics develop) and the estimated probability of extinction (*epslon*~t = 1 - *phi*~t) and origination (*gamma*) in the next interval if the previous interval *t-1* is either occupied or not by the genus *g*. The model design looks like:


```{r comment='', echo = F}
cat(readLines('dyn_model_vectorized.txt'), sep = '\n')
``` 


Besides allowing the estimation of biological parameters depicting taxon dynamics over time, state-space models enable the estimation of parameters depicting the probability of detecting a genus *p~[g]* in any geological formation and interval conditional on the genus realized occurrence *z~[t,g]*. The new detection/detection dataset (or the data that should be observed given the realization of the biological and observational processes) of genus *g* in interval *t* and geological formation *k*, denoted by the latent variable *y~[g,t,k]*, is another parameter estimated by the state-space model. The  *y~[g,t,k]* is the realization of a Bernoulli process which depends on *muY~[g,t,k]*, i.e. the probability of detecting a genus conditional on its realized occurrence (i.e., *z~[t,g] x p~[g]*). In this model, *p~[g]* only varies across *g* to *G* genera. We estimated genus detection probability across geological formations as each one of them represent a fieldwork effort to obtain fossil data, and might represent differential physical processes causing sedimentation/erosion, transport, deformation, and lost of paleontological/taphonomic information/affect fosssil preservation.
The input data consist of detections and non-detections (coded as 1 and 0, respectively) of each genus in each geological formation and interval. A genus was 'detected' if it appeared at least once in a geological formation and interval, and it was 'not-detected' if it not appeared in formations and intervals with at least one detected genus. Missing observations (coded as 'NA') comprise intervals and formations with no detected genus, and were removed from the state-space model. Nonetheless, the model estimated data for these missing observations using the existing data and the prior information (see below "Modeling approach").

## Derived parameters
Using the state-space model structure we were able to estimate parameters often used to make inference in paleontology, namely 'extinction probability', 'origination probability', 'proportional change', and 'equilibrium occupancy'. The first ones depict the probability of genus extinction and origination from t to t+1 (i.e., between each pair of adjacent time intervals). There were 16 probability estimates from Olenekian (early Triassic) to Tithonian (late Jurassic), as there is no transition from Olenekian to previous time (Changhsingiano, late Permian). We choose not to include Permian data in our state-space model because it would include a large taxonomic heterogeneity in our data, as the Permian-Triassic boundary (~252 Ma) was characterized by the most massive extinction on Earth's history (Benton 1995).

The proportional change (Bambach et al. 2004) depicts the difference in the number of taxa in the end relative to the beginning of a given period, as follows:

*PC[t]* = sum (z[t-1,]) - sum(z[t,]))/sum(z[t-1,])

Finally, the equilibrium occupancy showing the trend of a genus *g* to either decline or increase over time was estimated as follows:

*psi.eq[g]* = mean(gamma[t:T,g])/(mean(gamma[t:T,g])+mean(1-epslon[t:T,g]))



## Modeling approach

We estimated the state-space model parameters using Bayesian inference. Bayesian inference is a statistical method that applies the Bayes Theorem to update prior information about parameter values with observed data to then estimate the posterior probability of parameters. The prior information, generally represented by statistical distributions, are updated by their integration with data and likelihood estimation across independent Monte Carlo Markov Chains (MCMC). Variation in parameter estimates across posterior distribution draws represents an appropriate measure of parameter uncertainty (Kruschke & Liddell, 2018). In this sense, the Credible Intervals (CI), built using the quantiles of the posterior distribution draws of each parameter, depict the interval where most posterior distribution draws are and, therefore, delimit the area in which we have large certainty of finding the true parameter value (Kruschke & Liddell, 2018) [COPY OF MY JBI PAPER].

The priors for *phi[t,g]*, *gamma[t,g]*, *psi[t=1,g]*, and *p[g]* were all assumed to come from a uniform distribution ranging from 0 to 1 (*phi[t,g]*, *gamma[t,g]*, *psi[t=1,g]*, *p[g]* ~ U(0,1)).

Our state-space model was run using three independent MCMC (using the Gibbs Sampling algorithm implemented in JAGS; Plummer, 2003), with 15,000 iterations and a warm up period of 7,000 iterations each chain. We retained 1000 post-warm up draws of each chain, resulting in 3000 posterior distribution draws of each model parameter which were used to make statistical inference and test our hypotheses.

# Results
The analyzed dataset comprises 1139 genera observed at 17 time intervals and 1769 geological formations. Thus, if all intervals and geological formations were sampled we would have 30.073 possible observations in this dataset, per taxon/genus. However, the dataset is quite sparse, with many missing observations, and had only 7.1 % (n=2.149 observations) out of the possible observations per taxon. 




